---
title: "Lab 02: Linear Models"
author: "Juan Steibel and Austin Putz"
date: "Feb 01, 2023"
format:
  html:
    embed-resources: true
    theme: cosmo 
    toc: true
    toc-location: left
    number-depth: 3
    html-math-method: katex
    code-fold: false
---

<br> <br>

## 1. OLS with R

### A. Regression on continuous variables

Let's obtain the OLS of the longley dataset using matrix algebra operations and LM.

```{r}

# load packages
library(datasets)
library(Matrix)

# load dataset
data(longley)

# head
head(longley)

# y vector
y <- longley$Employed
y

# create X
X <- cbind(1, longley$GNP)
X

# rank of X
rankMatrix(X)

# X'X
XX <- t(X) %*% X
XX

# (X'X)^-1
XXi <- solve(XX)
XXi

# X'y
Xy <- t(X) %*% y

# b-hat
beta_hat <- XXi %*% Xy

# residuals (y - Xb)
ehat <- y - (X %*% beta_hat)

# residual variance (estimate)
sigma2_e_hat <- sum(ehat^2) / (nrow(X)-ncol(X))
sigma2_e_hat

# residual standard error (RMSE?)
sqrt(sigma2_e_hat)

# (X'X)^-1 * sigma^2_e
var_hat_beta_hat <- sigma2_e_hat*XXi
var_hat_beta_hat

# standard errors of coefficients
sqrt(diag(var_hat_beta_hat))

```

Now we can use $lm$ and compare the results

```{r}

# fit model
srm <- lm(Employed ~ GNP, data=longley)

# rank of X
srm$rank

# estimated means
srm$fitted.values

# y-hat
X %*% beta_hat

# easier comparison
cbind(
  srm$fitted.values,
  predict(srm),
  X %*% beta_hat
)

# residuals
cbind(
  residuals(srm),
  ehat
)

# summary gives a lot of details
summary(srm)

# but for obtaining regression coefficients, coef(srm is enough)
coef(srm)
beta_hat

# error standard deviation is available from summary method
summary(srm)$sigma
sqrt(sigma2_e_hat)

# estimated variance-covariance of coefficients
vcov(srm)
var_hat_beta_hat

# X'X also available
summary(srm)$cov.unscaled
XXi

# QR decomposition
srm$qr

# QR functions:
QR <- qr(X)
QR

Q1 = qr.Q(QR)
Q1
R1 = qr.R(QR)
R1

# solve for coefficients by hand
solve(R1) %*% t(Q1) %*% y 

# solve for coefficients with QR decomposition
cf <- qr.coef(QR, y)
cf

cbind(
  cf,
  beta_hat,
  coef(srm)
)

```

<br>

### B. Scaled predictors

We mentioned in class that this X matrix may present numerical problems due to very different scales for its columns (regressors) we can create a new variable by scaling the columns, but the intercept can't be scaled.

```{r}

# scale X
scale(X)

#  scale predictor
GNP_s <- scale(longley$GNP)
GNP_s

# check scale
round(mean(GNP_s), 5)
var(GNP_s)

```

See Homework for an exercise with scaled X.

<br> <br> <br>

## 2. Computing OLS

In class we learned about the OLS equations $(\boldsymbol{X}^{'}\boldsymbol{X})^{-1}\boldsymbol{X}^{'}\boldsymbol{y}$ and two methods for solving this system: 1) using direct inverse, 2) using the QR decomposition.

In this lab we will compare the two method using simulated data of different sizes according to the following parameters:

```{r}

# simulate (naive) data for regression analyses
N <- c(100,500,1000,10000) # number of samples
p <- c(10,50,100,500) # number of regressors
# all variables uniformly distributed

#a function to simulate data:
naive_reg<-function(N,p){
  X <- matrix(runif(N*p), N, p)
  y <- matrix(runif(N), N, 1)
  return(list(X=X, y=y))
}

# test
s1 <- naive_reg(100,10)
dim(s1$X)
length(s1$y)

```

Now we can test this using the microbenchmark package for the direct inverse and for the QR decomposition

```{r}
library(microbenchmark)
#continue :)

```

Follow class instructions to pursue some selected combinations and report back to the class and in your homework.

<br> <br> <br>

## 3. Building incidence matrices and solving OLS equations.

The function $model.matrix$ in R can be used to generate incidence matrices.

Use the dataset `npk`. Give a brief explanation of the dataset. Use the model.matrix function to generate incidence matrices for the following models and parameterizations:\

1.  $yield=N+P+K+e$ using corner parameterization.

2.  $yield=block+N+P+K+e$ using corner parameterization

3.  $yield=N+P+K+N \times P+N \times K+K \times P+N \times P \times K+e$ using corner parameterization

4.  $yield=N+P+K+N \times P+N \times K+K \times P+N \times P \times K+e$ using cell mean model

For each model present the incidence matrix $\boldsymbol{X}$, mark the columns corresponding to each factor or model terms, explain how the columns of $\boldsymbol{X}$ were selected to make the matrix full column rank.

```{r include=FALSE}

write_matex2 <- function(x) {
  options(digits=2)
  begin <- "\\begin{bmatrix}"
  end <- "\\end{bmatrix}"
  X <-
    apply(x, 1, function(x) {
      paste(
        paste(x, collapse = "&"),
        "\\\\"
      )
    })
  paste(c(begin, X, end), collapse = "")
}

```

```{r}

# data
data(npk)

# print first lines
head(npk)

# print model (design) matrix
mm1 <- model.matrix(yield~N+P+K,data=npk)

# print
print(mm1)
  
```

$$
`r write_matex2(mm1)`
$$ Where:\
column 1 = intercept,\
column 2 = N,\
column 3 = P,\
column 4 = K,

<br> <br> <br>

## 4. fitting linear models to a real dataset

Work with a dataset from Gualdrón et al. (2014). (doi: 10.1186/1471-2105-15-246)

```{r}

# set working directory
setwd("~/Documents/ISU/Classes/AnS_562/2023/Part_A/R/")

# read in data
dts <- read.table("pheno_ready.txt")

# print first rows
head(dts)

# write data
# write.table(
#   x = dts,
#   file = "swine_data.csv",
#   quote = FALSE,
#   sep = ",",
#   row.names = FALSE,
#   col.names = TRUE
# )

```

The columns are: ID: wt_birth: weight at birth car_wt: carcass weight num_ribs: number of ribs car_bf10: back fat depth at the 10th rib car_lma: loin muscle area sex: Barrow vs gilt per_duroc: estimated genomic proportion inherited from Duroc breed sire dam litter slgdt_cd: slaughtering group

-   Select a trait to work with among these:

    -   wt_birth: weight at birth

    -   car_wt: carcass weight

    -   num_ribs: number of ribs

    -   car_bf10: back fat depth at the 10th rib car\_

    -   lma: loin muscle area

-   are there any records that need to be excluded?

-   Discuss with the class/group which covariates/cofactors should be used for modeling

-   Once the candidates are selected, do a brief summary of each of those

-   Pursue variable/model selection using the $lm$ function and related tools

-   Extract incidence matrices and explain each column (hint: only present a few rows if necessary)

<br> <br> <br>























